pdp: An R Package for Creating Partial Dependence Plots

### 2016-09-13 manuscript/greenwell.tex
minor edits

It is not wise to draw conclusions from PDPs in regions outside the range of the training data. Here we describe two ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls. Rug displays are one-dimensional plots added to the axes. Both \code{partial} and \code{plotPartial} have a \code{rug} option that, when set to \code{TRUE}, will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal and vertical axes. The following snippet of code produces the left display in Figure~\ref{fig:partial_extrap}.
\begin{example}
partial(boston.rf, pred.var = "lstat", rug = TRUE, plot = TRUE)
\end{example}

In two or more dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on. When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is added to the plot; alternatively, you can set \code{chull = TRUE} in the call to \code{partial}, in which case only the region within the convex hull of the first two variables is plotted. Over interpreting the PDP outside of this region can be dangerous. The right display in Figure~\ref{fig:partial_extrap} was produced using:
\begin{example}
partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE, plot = TRUE)
\end{example}

@@

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of median home value using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points ban be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a specific set of values (e.g., quantiles of interest) can be supplied through the \code{pred.grid} argument. To illustrate, the following snippet of code computes the partial dependence of median home value on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
partial(boston.rf, "rm", plot = TRUE)
partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
\end{example}

@@

Next, we use MARS to model ozone concentration as a nonlinear function of the eight meteorological variables plus day of the year; we allow for up to three-way interactions.
\begin{example}
ozone.earth <- earth(ozone ~ ., data = ozone, degree = 3)
summary(ozone.earth)
\end{example}
The model produced a generalized $R^2$ of $0.79$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
\begin{itemize}
  \item \code{wind}: wind speed (mph) at Los Angeles International Airport (LAX)
  \item \code{temp}: temperature ($^oF$) at Sandburg Air Force Base
  \item \code{dpg}: the pressure gradient (mm Hg) from LAX to Dagget, CA
\end{itemize}
To understand this interaction, we can use a partial plot. However, since the partial dependece between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.

@@

Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded. This is illustrated in following snippet of code which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The result is displayed in Figure~\ref{fig:partial_par}.
\begin{example}
partial(ozone.earth, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, .parallel = TRUE)
\end{example}

@@

It is important to note that when using more than two predictor variables, \code{plotPartial} produces a trellis display. The first two variables given to \code{pred.var} are used for the horizontal and vertical axes, and additional variables define the panels. If the panel variables are continuous, then shingles\footnote{A shingle is a special Trellis data structure that consists of a numeric vector along with intervals that define the "levels" of the shingle. The intervals may be allowed to overlap.} are produced first using the equal count algorithm (see, for example, \code{?lattice::equal.count}). Hence, it is probably better to use categorical variables to define the panels in higher dimensional displays when possible, as was done in the previous example.

@@

To illustrate, consider Edgar Anderson's iris data from the \pkg{datasets} package. The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris:  setosa, versicolor, and virginica. We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} (the model parameters were determined using 5-fold cross-validation).
\begin{example}
library(e1071)
iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
                cost = 0.25, probability = TRUE)
\end{example}
\strong{Note:} the \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set the \code{probability = TRUE} option in the call to \code{svm}.

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}. Note that we had to supply the original training data to \code{partial} through the \code{train} option. This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, grid.resolution = 101, .progress = "text")
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}

@@

To round out our discussion, we provide one last example using a recently popular (and successful!) machine learning tool. XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable. The associated R package \pkg{xgboost} has been used to win a number of \href{https://www.kaggle.com/}{Kaggle competitions}. It has been shown to be many times faster than the well known \pkg{gbm} package. However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for displaying PDPs. Fortunately, the \pkg{pdp} package can be used to fill this gap.

@@

In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package. We also briefly discussed related approaches availble in other R packages as well. Suggestions to avoid extrpolation and high excecution times were discussed and demonstrated via examples. There is a bright future for \pkg{pdp} in terms of growth. For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response.

>>>>

It is not wise to draw conclusions from PDPs in regions outside the area of the training data. Here we describe two ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls. Rug displays are one-dimensional plots added to the axes. Both \code{partial} and \code{plotPartial} have a \code{rug} option that, when set to \code{TRUE}, will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal and vertical axes. The following snippet of code produces the left display in Figure~\ref{fig:partial_extrap}.
\begin{example}
partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE)
\end{example}

In two or more dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on. When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is computed; for example, if you set \code{chull = TRUE} in the call to \code{partial} only the region within the convex hull of the first two variables is plotted. Over interpreting the PDP outside of this region can be dangerous. The right display in Figure~\ref{fig:partial_extrap} was produced using:
\begin{example}
partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
\end{example}

@@

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of median home value using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To illustrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
partial(boston.rf, "rm", plot = TRUE)
partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
\end{example}

@@

Next, we use MARS to model ozone concentration as a nonlinear function of the eight meteorological variables plus day of the year; we allow for up to three-way interactions.
\begin{example}
ozone.mars <- earth(ozone ~ ., data = ozone, degree = 3)
summary(ozone.mars)
\end{example}
The model produced a generalized $R^2$ of $0.79$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
\begin{itemize}
  \item \code{wind}: wind speed (mph) at Los Angeles International Airport (LAX)
  \item \code{temp}: temperature ($^oF$) at Sandburg Air Force Base
  \item \code{dpg}: the pressure gradient (mm Hg) from LAX to Dagget, CA
\end{itemize}
To understand this interaction, we can use a PDP. However, since the partial dependece between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.

@@

Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded. This is illustrated in code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The result is displayed in Figure~\ref{fig:partial_par}.
\begin{example}
partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, .parallel = TRUE)
\end{example}

@@

It is important to note that when using more than two predictor variables, \code{plotPartial} produces a trellis display. The first two variables given to \code{pred.var} are used for the horizontal and vertical axes, and additional variables define the panels. If the panel variables are continuous, then shingles\footnote{A shingle is a special Trellis data structure that consists of a numeric vector along with intervals that define the "levels" of the shingle. The intervals may be allowed to overlap.} are produced first using the equal count algorithm (see, for example, \code{?lattice::equal.count}). Hence, it is probably better to use categorical variables to define the panels in higher dimensional displays when possible.

@@

To illustrate, we consider Edgar Anderson's iris data from the \pkg{datasets} package. The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris:  setosa, versicolor, and virginica. We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} (the tuning parameters were determined using 5-fold cross-validation).
\begin{example}
library(e1071)
iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
                cost = 0.25, probability = TRUE)
\end{example}
\strong{Note:} the \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set \code{probability = TRUE} in the call to \code{svm}.

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}. %Note that we had to supply the original training data to \code{partial} through the \code{train} option. This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, grid.resolution = 101, .progress = "text")
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}

@@

To round out our discussion, we provide one last example using a recently popular (and successful!) machine learning tool. XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable. The associated R package \pkg{xgboost} has been used to win a number of \href{https://www.kaggle.com/}{Kaggle competitions}. It has been shown to be many times faster than the well known \pkg{gbm} package. However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for constructing PDPs. Fortunately, the \pkg{pdp} package can be used to fill this gap.

@@

In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package. We also briefly discussed related approaches availble in other R packages as well. Suggestions to avoid extrpolation and high excecution times were discussed and demonstrated via examples. In terms of future development, \pkg{pdp} can be expanded in a number of ways. For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response.
